---
title: "Common Extensions"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=T, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```

```{r catchup}
# if needed
library(tidyverse)
library(lme4)

load('data/gpa.RData')

# if you want to run all the code, you'll need to install the following
# install.packages(c('sjstats', 'merTools', 'broom'))
```


## Additional Grouping Structure

### Crossed random effects

```{r load_pupils_data}
load('data/pupils.RData')

pupils
```

#### Primary and Secondary School Random Effects


```{r cross_classified}
pupils_crossed = lmer(achievement ~ sex + ses 
                      + (1|primary_school_id) + (1|secondary_school_id), 
                      data = pupils)

summary(pupils_crossed, correlation=F)
```

```{r crossed_vc}
VarCorr(pupils_crossed) %>% 
  print(comp=c('Var', 'Std'), digits=3)
```

```{r crossed_icc}
sjstats::icc(pupils_crossed) # relative proportion due to each school
```

```{r crossed_re}
str(ranef(pupils_crossed))
```


```{r crossed_re_plot, echo=FALSE}
merTools::plotREsim(merTools::REsim(pupils_crossed)) +
  theme_minimal()
```


### Hierarchical Structure

```{r nurses_data, echo=1}
load('data/nurses.RData')

nurses
```


```{r hierarchical}
nurses_hierarchical = lmer(stress ~ age + sex + experience 
                           + treatment + wardtype + hospsize 
                           + (1|hospital) + (1|hospital:ward), 
                           data = nurses)

nurses_hierarchical = lmer(stress ~ age  + sex + experience 
                           + treatment + wardtype + hospsize 
                           + (1|hospital/ward), 
                           data = nurses) # same thing!

summary(nurses_hierarchical, correlation=F)
```

```{r hierarchical_fixed}
nurses_hierarchical %>% 
  broom::tidy('fixed', conf.int=T) %>% 
  mutate_if(is.numeric, round, digits = 2)
```


As far as the fixed effects go, about the only thing that doesn't have a statistical effect is ward type.  


```{r hierarchical_random}
VarCorr(nurses_hierarchical)
```


### Crossed vs. Nested

### Crossed vs. Nested

The following shows the difference in the results from treating ward as a nested (within hospital) vs. crossed random effect. What do you notice is different?

```{r crossed_vs_nested}
nurses_hierarchical1 = lmer(stress ~ age  + sex + experience 
                           + treatment + wardtype + hospsize 
                           + (1|hospital) + (1|hospital:ward), data = nurses)

nurses_crossed1 = lmer(stress ~ age  + sex + experience 
                           + treatment + wardtype + hospsize 
                           + (1|hospital) + (1|ward), data = nurses)

nurses_crossed2 = lmer(stress ~ age  + sex + experience 
                           + treatment + wardtype + hospsize 
                           + (1|hospital) + (1|wardid), data = nurses)
```


```{r crossed_vs_nested}
VarCorr(nurses_hierarchical1)
```

```{r crossed_vs_crossed1}
VarCorr(nurses_crossed1)
```

```{r crossed_vs_crossed2}
VarCorr(nurses_crossed2)
```


The first hierarchical model and the second crossed version are identical. The second crossed is incorrectly labeled to be using a crossed notation for the model, as the label doesn't distinguish ward 1 in hospital 1 from ward 1 in hospital 2.  In the second crossed model, we use `wardid` instead of `ward`, so each ward has a unique id, and the proper structure is accounted for with  the crossed syntax.


## Residual Structure


### Heterogeneous Variances

```{r heterovar}
library(nlme)
heterovar_res = lme(gpa ~ occasion, 
                    data = gpa,
                    random = ~1|student, 
                    weights = varIdent(form = ~1|occasion))

summary(heterovar_res)
```

```{r relative_variances}
# values are relative to redisual variance and on the standard deviation scale
summary(heterovar_res$modelStruct)
```

```{r heterovar_glmmTMB}
library(glmmTMB)
heterovar_res2 = glmmTMB(gpa ~ occasion + (1|student) + diag(0 + occas |student), 
                         data = gpa)

summary(heterovar_res2) # you can ignore the Corr part in the random effects output
```

```{r glmmtmb_extract_variances}
vc_glmmtmb = VarCorr(heterovar_res2)
vc_glmmtmb = attr(vc_glmmtmb$cond$student.1, 'stddev')^2 + sigma(heterovar_res2)^2
vc_glmmtmb
```

### Autocorrelation

```{r corr_residual}
corr_res = lme(gpa ~ occasion, 
               data = gpa,
               random = ~1|student, 
               correlation = corAR1(form = ~occasion))

summary(corr_res)
```