---
title: "Mixed Models: Basics"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=T, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```


## Example: Student GPA

```{r basic_packages}
library(tidyverse)
```


```{r load_gpa_data}
load('data/gpa.RData')
gpa
```

## Application

### Standard Regression

A standard linear regression model. We have coefficients for the intercept and the effect of time, and in addition, the variance of the observations (residual standard error).

```{r gpa_lm}
gpa_lm = lm(gpa ~ occasion, data=gpa)
summary(gpa_lm)
```

### Mixed Model

```{r gpa_mixed}
library(lme4)
gpa_mixed = lmer(gpa ~ occasion + (1|student), data=gpa)
summary(gpa_mixed)
```

[As a test, replace `1|student` with `1|sample(1:10, 1200, replace = T)`.  As your variance due to arbitrary grouping is essentially 0, the residual error estimate is similar to the `lm` model.]


People always ask where the p-values are, but the answer is... complicated.  You can get interval estimates easily enough, even though they are poorly named for the variance components.

```{r gpa_mixed_confint}
confint(gpa_mixed)
```

#### Estimated Random Effects

Examine the random effects.

```{r gpa_mixed_ranef}
ranef(gpa_mixed)$student 
```

```{r gpa_mixed_rancoef}
coef(gpa_mixed)$student 
```

As we didn't allow the occasion effect to vary, it is constant.

#### Prediction

```{r gpa_mixed_prediction}
predict(gpa_mixed, re.form=NA) %>% head
```

## Adding a Cluster-level Covariate

See exercises.