---
title: "More Random Effects"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


```{r chunk_setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = T, message=F, warning=F, comment=NA, autodep=F, 
                      eval=T, cache.rebuild=F, cache=T, R.options=list(width=120), 
                      fig.width=8, fig.align = 'center', dev.args=list(bg = 'transparent'), dev='svglite')
```

```{r catchup}
# if needed
library(tidyverse)
library(lme4)

load('data/gpa.RData')

gpa_lm = lm(gpa ~ occasion, data=gpa)
```

## Application

Add a random slope to our previous example and examine the results.

```{r random_slope}
gpa_mixed =  lmer(gpa ~ occasion + (1 + occasion|student), data=gpa)
summary(gpa_mixed)
```

### Explore Random Effects

As before, we can examine the per-student random effects. 

```{r random_effects}
ranef(gpa_mixed)$student
```

Unlike before, we see each student's occasion effect.

```{r random_coefficients}
coef(gpa_mixed)$student
```


## Comparison to Many Regressions

The following code calculates regression models for each student (i.e. six observations apiece). Density plots are shown comparing regressions run for each student and the mixed model.

```{r by_group}
gpa_lm_by_group0 = lmList(gpa ~ occasion | student, gpa)
gpa_lm_by_group  = coef(gpa_lm_by_group0)

gint = 
  data_frame(Mixed=coef(gpa_mixed)$student[,1], Separate=gpa_lm_by_group[,1]) %>% 
  gather(key=Model, value=Intercept) %>% 
  ggplot(aes(x=Intercept)) +
  geom_density(aes(color=Model, fill=Model), alpha=.25) +
  scale_color_viridis_d(begin = .25, end = .75) +
  scale_fill_viridis_d(begin = .25, end = .75) +
  ggtitle('Intercepts') +
  labs(x='', y='') +
  xlim(c(1.5,4)) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.key.size=unit(2, 'mm'),
    legend.title=element_text(size=8),
    legend.text=element_text(size=8),
    legend.box.spacing=unit(0, 'in'),
    legend.position=c(.85,.75)
    )

gslopes = 
  data_frame(Mixed=coef(gpa_mixed)$student[,2], Separate=gpa_lm_by_group[,2]) %>% 
  gather(key=Model, value=Occasion) %>% 
  ggplot(aes(x=Occasion)) +
  geom_density(aes(color=Model, fill=Model), alpha=.25, show.legend=F) +
  scale_color_viridis_d(begin = .25, end = .75) +
  scale_fill_viridis_d(begin = .25, end = .75) +
  ggtitle('Slopes for occasion') +
  labs(x='', y='') +
  xlim(c(-.2,.4)) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )


library(patchwork)
gint + gslopes
```

## Visualization of Effects


Let's look at what the results are in terms of prediction. First we can look at the mixed effects results.

```{r visualize_mixed_fit}
# add if you want, not displayed as there are only a couple negative slopes
going_down = factor(rep(coef(gpa_mixed)$student[,'occasion']<0, e=6), labels=c('Up', 'Down'))

library(modelr) # allows us to add predictions to the data frame
gpa %>% 
  add_predictions(gpa_lm, var='lm') %>% 
  add_predictions(gpa_mixed, var='mixed') %>% 
  ggplot() + 
  geom_line(aes(x=occasion, y=mixed, group=student), alpha=.1, color='#00aaff') +
  geom_line(aes(x=occasion, y=lm, group=student), color='#ff5500') +
  labs(y='gpa') + 
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

With that in mind, now we can see the messier 'by-group' approach.  The general trend is ignored, and many more students are predicted with downward trends when they probably shouldn't be.

```{r visualize_by_group_fit}
going_down = factor(rep(gpa_lm_by_group[,'occasion']<0, e=6), labels=c('Upward', 'Downward'))

gpa %>% 
  mutate(stufit=fitted(gpa_lm_by_group0)) %>% 
  add_predictions(gpa_lm, var='gpa') %>% 
  add_predictions(gpa_lm, var='lm') %>% 
  ggplot() +
  geom_line(aes(x=occasion, y=stufit, group=student, color=going_down, alpha = going_down)) +
  geom_line(aes(x=occasion, y=lm), 
            color='#ff5500') +
  labs(y='gpa') + 
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```


